\section{CASS: A Generic Curry Analysis Server System}
\label{sec-cass}

CASS\index{CASS}\index{analyzing programs}\index{program!analysis}
(Curry Analysis Server System)
is a tool for the analysis of Curry programs.
CASS is generic so that various kinds of analyses (e.g., groundness,
non-determinism, demanded arguments) can be easily integrated into CASS.
In order to analyze larger applications consisting of dozens or hundreds
of modules, CASS supports a modular and incremental analysis of
programs. Moreover, it can be used by different programming tools,
like documentation generators, analysis environments, program
optimizers, as well as Eclipse-based development environments. For
this purpose, CASS can also be invoked as a server system to get a
language-independent access to its functionality. CASS is completely
implemented Curry as a master/worker architecture to exploit parallel
or distributed execution environments.
The general design and architecture of CASS is described
in \cite{HanusSkrlac14}.
In the following, CASS is presented from a perspective
of a programmer who is interested to analyze Curry programs.

\subsection{Installation}

The current implementation of CASS is a package
managed by the Curry Package Manager CPM.
Thus, to install the newest version of CASS, use the following commands:
%
\begin{curry}
> cypm update
> cypm install cass
\end{curry}
%
This downloads the newest package, compiles it, and places
the executable \code{cass} into the directory \code{\$HOME/.cpm/bin}.
Hence it is recommended to add this directory to your path
in order to execute CASS as described below.

\subsection{Using CASS to Analyze Programs}

CASS is intended to analyze various operational properties
of Curry programs. Currently, it contains more than a dozen
program analyses for various properties.
Since most of these analyses are based on abstract interpretations,
they usually approximate program properties.
To see the list of all available analyses, use the help option of CASS:
\begin{curry}
> cass -h
Usage: $\ldots$
$\vdots$
Registered analyses names:
$\ldots$
Demand          : Demanded arguments
Deterministic   : Deterministic operations
$\vdots$
\end{curry}
More information about the meaning of the various analyses
can be obtained by adding the short name of the analysis:
\begin{curry}
> cass -h Deterministic
$\ldots$
\end{curry}
For instance, consider the following Curry module \code{Rev.curry}:
\begin{curry}
append :: [a] -> [a] -> [a]
append []     ys = ys
append (x:xs) ys = x : append xs ys

rev :: [a] -> [a]
rev []     = []
rev (x:xs) = append (rev xs) [x]

main :: Int -> Int -> [Int]
main x y = rev [x .. y]
\end{curry}
%
CASS supports three different usage modes to analyze this program.

\subsubsection{Batch Mode}

In the batch mode, CASS is started as a separate application
via the shell command \code{cass},
where the analysis name and the name of the module to be analyzed
must be provided:\footnote{More output is generated when
the property \code{debugLevel} is changed in the configuration file
\code{.curryanalysisrc} which is installed in the user's home directory
when CASS is started for the first time.}
\begin{curry}
> cass Demand Rev
append : demanded arguments: 1
main : demanded arguments: 1,2
rev : demanded arguments: 1
\end{curry}
The \code{Demand} analysis shows the list of argument positions
(e.g., 1 for the first argument) which are demanded in order
to reduce an application of the operation to some constructor-rooted value.
Here we can see that both arguments of \code{main} are demanded
whereas only the first argument of \code{append} is demanded.
This information could be used in a Curry compiler
to produce more efficient target code.

The batch mode is useful to test a new analysis and get the information
in human-readable form so that one can experiment
with different abstractions or analysis methods.

\subsubsection{API Mode}

The API mode is intended to use analysis information in some
application implemented in Curry. Since CASS is implemented in Curry,
one can import the modules of the CASS implementation and
use the CASS interface operations to start an analysis and use the
computed results. For instance, CASS provides an operation
(defined in the module \code{CASS.Server})
\begin{curry}
analyzeGeneric :: Analysis a -> String -> IO (Either (ProgInfo a) String)
\end{curry}
to apply an analysis (first argument) to some module (whose name is
given in the second argument). The result is either the analysis
information computed for this module or an error message in case of
some execution error.

In order to use CASS via the API mode in a Curry program,
one has to use the package \code{cass} by the Curry package manager CPM
(the subsequent explanation assumes familiarity with the basic
features of CPM):
\begin{enumerate}
\item
Add the dependency on package \code{cass} and also on
package \code{cass-analysis}, which contains some base definitions,
in the package specification file \code{package.json}.
\item
Install these dependencies by \ccode{cypm install}.
\end{enumerate}
Then you can import in your application the modules
provided by CASS.

The module \code{Analysis.ProgInfo} (from package \code{cass-analysis})
contains operations to access the analysis information computed by CASS.
For instance, the operation
\begin{curry}
lookupProgInfo:: QName -> ProgInfo a -> Maybe a
\end{curry}
returns the information about a given qualified name in the
analysis information, if it exists.
As a simple example, consider the demand analysis which is implemented
in the module \code{Analysis.Demandedness} by the following operation:
\begin{curry}
demandAnalysis :: Analysis DemandedArgs
\end{curry}
\code{DemendedArgs} is just a type synonym for \code{[Int]}.
We can use this analysis in the following simple program:
\begin{currynomath}
import CASS.Server           ( analyzeGeneric )
import Analysis.ProgInfo     ( lookupProgInfo )
import Analysis.Demandedness ( demandAnalysis )

demandedArgumentsOf :: String -> String -> IO [Int]
demandedArgumentsOf modname fname = do
  deminfo <- analyzeGeneric demandAnalysis modname >>= return . either id error
  return $ maybe [] id (lookupProgInfo (modname,fname) deminfo)
\end{currynomath} %$
Of course, in a realistic program, the program analysis
is performed only once and the computed information \code{deminfo}
is passed around to access it several times.
Nevertheless, we can use this simple program to compute the
demanded arguments of \code{Rev.main}:
\begin{curry}
$\ldots$> demandedArgumentsOf "Rev" "main"
[1,2]
\end{curry}


\subsubsection{Server Mode}

The server mode of CASS can be used in an application implemented in
some language that does not have a direct interface to Curry.
In this case, one can connect to CASS via
some socket using a simple communication protocol that is specified
in the file \code{Protocol.txt} (in package \code{cass})
and sketched below.

To start CASS in the server mode, one has to execute the command
\begin{curry}
> cass --server [ -p <port> ]
\end{curry}
where an optional port number for the communication can be
provided. Otherwise, a free port number is chosen and shown. In the
server mode, CASS understands the following commands:
\begin{curry}
GetAnalysis
SetCurryPath <dir1>:<dir2>:...
AnalyzeModule          <analysis name> <output type> <module name>
AnalyzeInterface       <analysis name> <output type> <module name>
AnalyzeFunction        <analysis name> <output type> <module name> <function name>
AnalyzeDataConstructor <analysis name> <output type> <module name> <constructor name>
AnalyzeTypeConstructor <analysis name> <output type> <module name> <type name>
StopServer
\end{curry}
The output type can be \code{Text}, \code{CurryTerm}, or \code{XML}.
The answer to each request can have two formats:
\begin{curry}
error <error message>
\end{curry}
if an execution error occured, or
\begin{curry}
ok <n>
<result text>
\end{curry}
where \code{<n>} is the number of lines of the result text.
For instance, the answer to the command \code{GetAnalysis}
is a list of all available analyses. The list has the form
\begin{curry}
<analysis name> <output type>
\end{curry}
For instance, a communication could be:
\begin{curry}
> GetAnalysis
< ok 5
< Deterministic CurryTerm
< Deterministic Text
< Deterministic XML
< HigherOrder   CurryTerm
< DependsOn     CurryTerm
\end{curry}
The command \code{SetCurryPath} instructs CASS to use the given
directories to search for modules to be analyzed. This is necessary
since the CASS server might be started in a different location than
its client.

Complete modules are analyzed by \code{AnalyzeModule}, whereas
\code{AnalyzeInterface} returns only the analysis information of exported
entities. Furthermore, the analysis results of individual functions,
data or type constructors are returned with the remaining analysis
commands. Finally, \code{StopServer} terminates the CASS server.

For instance, if we start CASS by
\begin{curry}
> cass --server -p 12345
\end{curry}
we can communicate with CASS as follows (user inputs are prefixed by \ccode{>});
\begin{curry}
> telnet localhost 12345
Connected to localhost.
> GetAnalysis
ok 57
Overlapping XML
Overlapping CurryTerm
Overlapping Text
Deterministic XML
...
> AnalyzeModule Demand Text Rev
ok 3
append : demanded arguments: 1
main : demanded arguments: 1,2
rev : demanded arguments: 1
> AnalyzeModule Demand CurryTerm Rev
ok 1
[(("Rev","append"),"demanded arguments: 1"),(("Rev","main"),"demanded arguments: 1,2"),(("Rev","rev"),"demanded arguments: 1")]
> AnalyzeModule Demand XML Rev
ok 19
<?xml version="1.0" standalone="yes"?>

<results>
  <operation>
    <module>Rev</module>
    <name>append</name>
    <result>demanded arguments: 1</result>
  </operation>
  <operation>
    <module>Rev</module>
    <name>main</name>
    <result>demanded arguments: 1,2</result>
  </operation>
  <operation>
    <module>Rev</module>
    <name>rev</name>
    <result>demanded arguments: 1</result>
  </operation>
</results>
> StopServer
ok 0
Connection closed by foreign host.
\end{curry}


\subsection{Implementing Program Analyses}

This section explains the implementation of program analyses
available in CASS.
Since CASS is implemented in Curry, a program analysis
must also be implemented in Curry and added to the source code of CASS.
Therefore, one has to download the source code
which is easily done by the command
\begin{curry}
> cypm checkout cass
\end{curry}
This downloads the most recent version of CASS as a Curry package
into the directory \code{cass}.

Each program analysis accessible by CASS must be registered
in the CASS module \code{CASS.Registry}.
Such an analysis must contain an operation of type
\begin{curry}
Analysis a
\end{curry}
where \ccode{a} denotes the type of analysis results.
Furthermore, the analysis must also contain a ``show'' operation of type
\begin{curry}
AOutFormat -> a -> String
\end{curry}
intended to show the analysis results in various formats.
The type \code{AOutFormat} is defined in module
\code{Analysis.Types} of package \code{cass-analysis} as
\begin{curry}
data AOutFormat = AText | ANote
\end{curry}
It is intended to specify the desired kind of output,
e.g., \code{AText} for a longer standard
textual representation or \code{ANote} for a short note
(e.g., in the Curry Browser).

Thus, in order to add a new analysis to CASS, one has to do the following
steps:
\begin{enumerate}
\item
Implement a corresponding analysis operation and show operation.
\item
Registering it in the module \code{CASS.Registry}
(in the constant \code{registeredAnalysis}).
\item
Compile/install the modified CASS implementation.
\end{enumerate}
In the following, we explain these steps by some examples.
%
For instance, the \code{Overlapping} analysis should indicate
whether a Curry operation is defined by overlapping rules.
This analysis can be implemented as a function
\begin{curry}
overlapAnalysis :: Analysis Bool
\end{curry}
so that the analysis result is \code{False} if the analyzed
operation is not defined by overlapping rules.

In general, an analysis is implemented as a mapping from Curry operations,
represented in FlatCurry, into the analysis result.
Hence, to implement the \code{Overlapping} analysis, we define
the following operation on function declarations in FlatCurry format:
\begin{curry}
import FlatCurry.Types
$\ldots$
isOverlappingFunction :: FuncDecl -> Bool
isOverlappingFunction (Func _ _ _ _ (Rule _ e))   = orInExpr e
isOverlappingFunction (Func f _ _ _ (External _)) = f == ("Prelude","?")

-- Check an expression for occurrences of Or:
orInExpr :: Expr -> Bool
orInExpr (Var _)       = False
orInExpr (Lit _)       = False
orInExpr (Comb _ f es) = f == ("Prelude","?") || any orInExpr es
orInExpr (Free _ e)    = orInExpr e
orInExpr (Let bs e)    = any orInExpr (map snd bs) || orInExpr e
orInExpr (Or _ _)      = True
orInExpr (Case _ e bs) = orInExpr e || any orInBranch bs
 where orInBranch (Branch _ be) = orInExpr be
orInExpr (Typed e _)   = orInExpr e
\end{curry}
%
In order to support the inclusion of different kinds of analyses in CASS,
CASS offers several constructor operations for the abstract type
\ccode{Analysis a}
(which is defined in module \code{Analysis.Types}).
Each analysis has a name provided as a first argument
to these constructors. The name is used to store the
analysis information persistently and to pass specific analysis tasks
to analysis workers.
For instance, a simple function analysis which depends only on a
given function definition can be defined by the
analysis constructor
\begin{curry}
simpleFuncAnalysis :: String -> (FuncDecl -> a) -> Analysis a
\end{curry}
The arguments are the analysis name and the actual analysis function.
Hence, the ``overlapping rules'' analysis can be specified as
\begin{curry}
import Analysis.Types
$\ldots$
overlapAnalysis :: Analysis Bool
overlapAnalysis = simpleFuncAnalysis "Overlapping" isOverlappingFunction
\end{curry}
In order to integrate this analysis into CASS, we also have to
define an operation to show the analysis results in a human-readable form:
%
\begin{curry}
showOverlap :: AOutFormat -> Bool -> String
showOverlap _     True  = "overlapping"
showOverlap AText False = "non-overlapping"
showOverlap ANote False = ""
\end{curry}
%
Here, the typical case of non-overlapping rules is not printed
in case of short notes.

Now we have all elements available in order to add this analysis to CASS.
To support this easily, there is an operation
%
\begin{curry}
cassAnalysis :: (Read a, Show a, Eq a)
             => String -> Analysis a -> (AOutFormat -> a -> String)
             -> RegisteredAnalysis
\end{curry}
%
to transform an analysis with some title, an analysis operation,
and a ``show'' operation into an analysis ready to be registered in CASS.
The actually registered analyses are specified by the constant
%
\begin{curry}
registeredAnalysis :: [RegisteredAnalysis]
\end{curry}
%
defined in module \code{CASS.Registry}.
Hence, the \code{Overlapping} can be integrated into CASS
by adding it to the definition of \code{registeredAnalysis}, e.g.,
%
\begin{curry}
registeredAnalysis :: [RegisteredAnalysis]
registeredAnalysis =
  [
  $\vdots$
  cassAnalysis "Overlapping rules" overlapAnalysis showOverlap
  $\vdots$
  ]
\end{curry}
%
As a final step, we have to compile and install this extended
version of CASS by executing
\begin{curry}
> cypm install
\end{curry}
in the downloaded package. After this step, one can executed
\begin{curry}
> cass --help
\end{curry}
to check whether the \code{Overlapping} analysis occurs in the list
of registered analyses names.

To show an example of a more complex kind of analysis, we consider
a determinism analysis.
Such an analysis could be based
on an abstract domain described by the data type
\begin{curry}
data Deterministic = NDet | Det
\end{curry}
Here, \code{Det} is interpreted as ``the operation always evaluates
in a deterministic manner on ground constructor terms.''
However, \code{NDet} is interpreted as ``the operation \emph{might}
evaluate in different ways for given ground constructor terms.''
The apparent imprecision is due to the approximation of the analysis.
For instance, if the function \code{f} is defined by overlapping rules
and the function \code{g} \emph{might} call \code{f}, then \code{g}
is judged as non-deterministic (since it is generally undecidable
whether \code{f} is actually called by \code{g} in some run of the
program).

The determinism analysis requires to examine the current
function as well as all directly or indirectly called functions
for overlapping rules.
Due to recursive function definitions, this analysis cannot be done
in one shot for a given function---it requires a fixpoint computation.
CASS provides such fixpoint computations and simplifies
its implementation by requiring only the implementation
of an operation of type
\begin{curry}
FuncDecl -> [(QName,a)] -> a
\end{curry}
where \ccode{a} denotes the type of abstract values.
The second argument of type \code{[(QName,a)]}
represents the currently known analysis values
for the functions \emph{directly} used in this function declaration.
Hence, in the implementation one can assume that the analysis
results of all functions occurring in the definition of
the function to be analyzed are already known, although they
will be approximated by a fixpoint computation performed by CASS.
Technically, the abstract values must be a domain with some
bottom element and the analysis operation must be monotone.
Since this is not checked by CASS, we omit these details.

In our example, the determinism analysis can be implemented
by the following operation:
\begin{curry}
detFunc :: FuncDecl -> [(QName,Deterministic)] -> Deterministic
detFunc (Func f _ _ _ (External _)) _           = f == ("Prelude","?")
detFunc (Func f _ _ _ (Rule _ e))   calledFuncs =
  if orInExpr e || freeVarInExpr e || any (==NDet) (map snd calledFuncs)
    then NDet
    else Det
\end{curry}
Thus, it computes the abstract value \code{NDet}
if the function itself is defined by overlapping rules or
contains free variables that might cause non-deterministic guessing
(we omit the definition of \code{freeVarInExpr} since it is quite
similar to \code{orInExpr}), or
if it depends on some non-deterministic function.

To support the integration of such fixpoint analyses in CASS,
there exists the following analysis constructor:
\begin{curry}
dependencyFuncAnalysis :: String -> a -> (FuncDecl -> [(QName,a)] -> a)
                       -> Analysis a
\end{curry}
Here, the second argument specifies the start value of the fixpoint analysis,
i.e., the bottom element of the abstract domain.
Hence, the complete determinism analysis can be specified as
%
\begin{curry}
detAnalysis :: Analysis Deterministic
detAnalysis = dependencyFuncAnalysis "Deterministic" Det detFunc
\end{curry}
%
In order to register this analysis, we define a show function
%
\begin{curry}
showDet :: AOutFormat -> Deterministic -> String
showDet _     NDet = "non-deterministic"
showDet AText Det  = "deterministic"
showDet ANote Det  = ""
\end{curry}
%
extend the definiton of \code{registeredAnalysis} by the line
%
\begin{curry}
cassAnalysis "Deterministic operations" detAnalysis showDet
\end{curry}
%
and compile and install the package.

This simple definition is sufficient to execute this analysis with CASS,
since the analysis system takes care of computing fixpoints,
calling the analysis functions with appropriate values,
analyzing imported modules, caching analysis results, etc.
The actual analysis time depends on the size of modules and their imports,
the size of the dependencies, and the
number of fixpoint iterations (which depends also on the depth
of the abstract domain).\footnote{CASS supports different
methods to compute fixpoints, see
the property \code{fixpoint} in the configuration file
\code{.curryanalysisrc} which is installed in the user's home directory
when CASS is started for the first time.
This property can also be set in the command to invoke CASS.}
Beyond the analysis time, it is also important that the analysis
terminates, which is not ensured in general fixpoint computations.
Termination can be achieved by using
an abstract domain with finitely many values and defining
the analysis function so that it is monotone w.r.t.\ some ordering
on the abstract values.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
