\documentclass[11pt,fleqn]{article}

\usepackage[T1]{fontenc}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

\setlength{\textwidth}{16.5cm}
\setlength{\textheight}{23cm}
\renewcommand{\baselinestretch}{1.1}
\setlength{\topmargin}{-1cm}
\setlength{\oddsidemargin}{0cm}
\setlength{\evensidemargin}{0cm}
\setlength{\marginparwidth}{0.0cm}
\setlength{\marginparsep}{0.0cm}

\newlength{\figurewidth}
\setlength{\figurewidth}{\textwidth}
\addtolength{\figurewidth}{-0.4cm}

% font for program texts
\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
\newcommand{\codefont}{\small\tt}

\usepackage{listings}
\lstset{aboveskip=1.5ex,
        belowskip=1.5ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        xleftmargin=2ex,
        basewidth=0.52em,
        basicstyle=\small\ttfamily}

\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
       }

\lstnewenvironment{curry}{}{}
\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
\newcommand{\listline}{\vrule width0pt depth1.75ex}

% program text in normal text
\newcommand{\code}[1]{\mbox{\codefont #1}}

% program text in normal text with apostrophs
\newcommand{\ccode}[1]{``\code{#1}''}

\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index

\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
\newcommand{\CYS}{Curry\xspace} % name of the Curry system described here
\newcommand{\cyshome}{\mbox{\textit{curryhome}}\xspace} % symbolic installation directory

\newcommand{\us}{\char95\xspace} % underscore

\begin{document}
\sloppy

\newcommand{\curryppdocs}{.}
\input{manual.tex}
\clearpage
\appendix
\input{sqlsyntax}

% Bibliography
\begin{thebibliography}{10}

\bibitem{AntoyHanus12PADL}
S.~Antoy and M.~Hanus.
\newblock Contracts and specifications for functional logic programming.
\newblock In {\em Proc. of the 14th International Symposium on Practical
  Aspects of Declarative Languages (PADL 2012)}, pages 33--47. Springer LNCS
  7149, 2012.

\bibitem{AntoyHanus14}
S.~Antoy and M.~Hanus.
\newblock Curry without {Success}.
\newblock In {\em Proc.\ of the 23rd International Workshop on Functional and
  (Constraint) Logic Programming (WFLP 2014)}, volume 1335 of {\em {CEUR}
  Workshop Proceedings}, pages 140--154. CEUR-WS.org, 2014.

\bibitem{AntoyHanus16PADL}
S.~Antoy and M.~Hanus.
\newblock Default rules for {Curry}.
\newblock In {\em Proc. of the 18th International Symposium on Practical
  Aspects of Declarative Languages (PADL 2016)}, pages 65--82. Springer LNCS
  9585, 2016.

\bibitem{BrasselHanusMueller08PADL}
B.~Bra{\ss}el, M.~Hanus, and M.~M{\"u}ller.
\newblock High-level database programming in {Curry}.
\newblock In {\em Proc. of the Tenth International Symposium on Practical
  Aspects of Declarative Languages (PADL'08)}, pages 316--332. Springer LNCS
  4902, 2008.

\bibitem{Hanus01PADL}
M.~Hanus.
\newblock High-level server side web scripting in {C}urry.
\newblock In {\em Proc.\ of the Third International Symposium on Practical
  Aspects of Declarative Languages (PADL'01)}, pages 76--92. Springer LNCS
  1990, 2001.

\bibitem{HanusKrone17EPTCS}
M.~Hanus and J.~Krone.
\newblock A typeful integration of {SQL} into {Curry}.
\newblock In {\em Proceedings of the 24th International Workshop on Functional
  and (Constraint) Logic Programming}, volume 234 of {\em Electronic
  Proceedings in Theoretical Computer Science}, pages 104--119. Open Publishing
  Association, 2017.

\bibitem{Krone15}
J.~Krone.
\newblock Integration of {SQL} into {Curry}.
\newblock Master's thesis, University of Kiel, 2015.

\end{thebibliography}

\end{document}
