\subsection {Datatypes with Field Labels}\label{flab}

A datatype declaration may optionally define data constructors
with field labels.\footnote{Field labels are quite similar
to Haskell \cite{PeytonJones03Haskell} so that we adopt most of
the description of Haskell here.}
These field labels can be used to
construct, select from, and update fields in a manner that is
independent of the overall structure of the datatype.

\subsubsection{Declaration of Constructors with Labeled Fields}

A data constructor of arity $n$ creates an object with $n$ components.
These components are normally accessed positionally as arguments to
the constructor in expressions or patterns.  For large datatypes it is
useful to assign \emph{field labels}\index{label}\index{field label}
to the components of a data
object. This allows a specific field to be referenced independently of
its location within the constructor.  A constructor definition in a
data declaration may assign labels to the fields of the constructor,
using the \emph{record syntax}\index{record syntax}\pindex{\{\ldots{}\}}
\code{C\,\,\{}\ldots\code{\}}.
Constructors using field labels
may be freely mixed with constructors without them. A constructor with
associated field labels may still be used as an ordinary constructor.
The various use of labels (see below)
are simply a shorthand for operations using an
underlying positional constructor. The arguments to the positional
constructor occur in the same order as the labeled fields.

\newcommand{\trans}[1]{$[\![$#1$]\!]$}

\translation{\\
\trans{$C$ \{ $lts$ \}} $=$ $C$ \trans{$lts$}\\
\trans{$lt$, $lts$} $=$ \trans{$lt$} \trans{$lts$}\\
\trans{$l$, $ls$\,::\,$t$} $=$ $t$ \trans{$ls$\,::\,$t$}\\
\trans{$l$\,::\,$t$} $=$ $t$
}

\noindent
For example, the definition using field labels
\begin{curry}
data Person = Person {firstName, lastName :: String, age :: Int}
            | Agent { firstName, lastName :: String, trueIdentity :: Person }
\end{curry}
is translated to
\begin{curry}
data Person = Person String String Int
            | Agent String String Person
\end{curry}
%
A data declaration may use the same field label in multiple
constructors as long as the typing of the field is the same in all
cases after type synonym expansion. A label cannot be shared by more
than one type in scope. Field names share the top-level name space with
ordinary definition of functions % and class methods
and must not conflict with other top-level names in scope.

Consider the following example:
\begin{curry}
data S = S1 { x :: Int } | S2 { x :: Int  } -- OK
data T = T1 { y :: Int } | T2 { y :: Bool } -- BAD
\end{curry}
Here S is legal but T is not, because y is given inconsistent typings in the latter.

\subsubsection{Field Selection}\label{flab-sel}
Field labels are used as \emph{selector functions},\index{selector function}
i.e., each field label serves as a function
that extracts the field from an object. Selectors are top-level
bindings and so they may be shadowed by local variables but cannot
conflict with other top-level bindings of the same name.  This
shadowing only affects selector functions; in record construction
(Section \ref{flab-constr}) and update (Section \ref{flab-upd}), field
labels cannot be confused with ordinary variables.

\translation{A field label $lab$ introduces a selector function defined as:\\
\code{$lab$ ($C_1$ $p_{11}$ \ldots $p_{1k_1}$) = x}\\
\ldots\\
\code{$lab$ ($C_n$ $p_{n1}$ \ldots $p_{nk_n}$) = x}\\
where $C_1\ldots C_n$ are all the constructors of the datatype containing
a field labeled with $lab$, $p_{ij}$ is \code{x} when $lab$ labels the $j$th
component of $C_i$ or \code{\us} otherwise.}

\noindent
For example the definition of \code{Person} above introduces the selector functions
\begin{curry}
firstName :: Person -> String
firstName (Person x _ _) = x
firstName (Agent  x _ _) = x

lastName :: Person -> String
lastName (Person _ x _) = x
lastName (Agent  _ x _) = x

age :: Person -> Int
age (Person _ _ x) = x

trueIdentity :: Person -> Person
trueIdentity (Agent  _ _ x) = x
\end{curry}

\subsubsection{Construction Using Field Labels}\label{flab-constr}

A constructor with labeled fields may be used to construct a value in
which the components are specified by name rather than by
position. In this case, the components are enclosed by braces.
Construction using field labels is subject to the following
constraints:
%
\begin{itemize}
\item Only field labels declared with the specified constructor may be
  mentioned.
\item A field label may not be mentioned more than once.
\item Fields not mentioned are initialized to different free variables.
\end{itemize}
%
The expression \code{C\{\}}, where \code{C} is a data constructor,
is legal \emph{whether or not  \code{C} was declared with record syntax};
it denotes \code{C Prelude.unknown$_1$ \ldots{} Prelude.unknown$_n$}
where $n$ is the arity of \code{C}.
Note that this will introduce the constructor \code{C} with
$n$ \emph{different} free variables as arguments.

\translation{In the binding $f = v$, the field $f$ labels $v$.\\
\centerline{\code{$C$ \{ $bs$ \} = $C$ ($pick_1^C\ bs\ \code{Prelude.unknown}$) \ldots{}
($pick_k^C\ bs\ \code{Prelude.unknown}$)}}
where $k$ is the arity of $C$.\\
The auxiliary function $pick_i^C\ bs\ d$ is defined as follows:

If the $i$th component of a constructor $C$ has the field label $f$ and $f = v$
appears in the binding list $bs$, then $pick_i^C\ bs\ d$ is $v$.
Otherwise, $pick_i^C\ bs\ d$ is the default value $d$.}

\noindent
For example, a \code{Person} can be constructed by
\begin{curry}
smith = Agent {lastName="Smith", firstName="Agent"}
\end{curry}
which is equivalent to the following agent, whose true identity
might be any person:
\begin{curry}
smith = Agent "Agent" "Smith" _
\end{curry}

\subsubsection{Updates Using Field Labels}\label{flab-upd}

Values belonging to a datatype with field labels may be
non-destructively updated. This creates a new value in which the
specified field values replace those in the existing value. Updates
are restricted in the following ways:

\begin{itemize}
\item All labels must be taken from the same datatype.
\item No label may be mentioned more than once.
\item The computation fails when the value being updated does not contain all of the specified
labels.
\end{itemize}

\translation{Using the prior definition of $pick$,\\
\code{$e$ \{ $bs$ \} $=$ let $upd$ ($C_1$ $v_1$\ldots{}$v_{k_1}$) = $C_1$ ($pick_1^{C_1}$ $bs$ $v_1$) \ldots{} ($pick_{k_1}^{C_1}$ $bs$ $v_{k_1}$)}\\
\hspace*{2.7cm}\ldots\\
\hspace*{2.7cm}\code{$upd$ ($C_j$ $v_1$\ldots{}$v_{k_j}$) = $C_j$ ($pick_1^{C_j}$ $bs$ $v_1$) \ldots{} ($pick_{k_j}^{C_j}$ $bs$ $v_{k_j}$)}\\
\hspace*{2.23cm}\code{in $upd~e$}\\[1ex]
where $\{C_1,\ldots,C_j\}$ is the set of constructors containing \emph{all} labels
in $bs$, $k_i$ is the arity of $C_i$, and $upd$ is a new function name.}

\noindent
For example, after watching a few more movies, we might want to update
our information about \code{smith}. We can do so by writing
\code{smith\{trueIdentity=complement neo\}}, which is equivalent to
\begin{curry}
let upd (Agent fn ln ti) = Agent (fn) (ln) (complement neo)
 in upd smith
\end{curry}

\subsubsection{Pattern Matching Using Field Labels}\label{flab-pat}

A constructor with labeled fields may be used to specify a pattern in
which the components are identified by name rather than by position.
Matching against a constructor using labeled fields is the same as
matching ordinary constructor patterns except that the fields are
matched in the order they are named in the field list. All listed
fields must be declared by the constructor; fields may not be named
more than once. Fields not named by the pattern are ignored (matched
against \code{\us}).

\translation{Using the prior definition of $pick$,\\
\code{$C$ \{ bs \} = ($C$ ($pick_1^C$ $bs$ _) \ldots{} ($pick_{k}^C$ $bs$ _))}\\
where $k$ is the arity of $C$.}

\noindent
For example, we could define a Smith-tester by writing:
\begin{curry}
isSmith Agent{lastName="Smith"} = success
\end{curry}
which is equivalent to
\begin{curry}
isSmith (Agent _ "Smith" _) = success
\end{curry}

\subsubsection{Field Labels and Modules}

As described in Curry report, there are two forms of
exporting a data type $T$: The simple name $T$ exports only the types name
without constructors, whereas $T$\code{(..)} also exports all
constructors. Analogously, the form $T$ does not export any field labels,
whereas $T$\code{(..)} exports all constructors and all field labels.

There is the additional possibility of exporting the selector functions which
are implicitly introduced by declaring a constructor with field labels,
(see Section~\ref{flab-sel}).
In contrast to \cite{PeytonJones03Haskell}, this does
not enable the possibility of updating a value using the field label syntax
(see Section~\ref{flab-upd}). This is prohibited mainly for two reasons:
\begin{itemize}
\item The translations given in this section are still valid in the context of modules.
\item From a software engineering point of view, it can make sense to
  export selectors for a certain type without allowing updates on it.
  Changing a given value is more often constrained by integrity
  properties than retrieving partial information from it.
\end{itemize}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual"
%%% End: 
