\documentclass[11pt,fleqn]{article}

\usepackage[T1]{fontenc}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

\setlength{\textwidth}{16.5cm}
\setlength{\textheight}{23cm}
\renewcommand{\baselinestretch}{1.1}
\setlength{\topmargin}{-1cm}
\setlength{\oddsidemargin}{0cm}
\setlength{\evensidemargin}{0cm}
\setlength{\marginparwidth}{0.0cm}
\setlength{\marginparsep}{0.0cm}

\newlength{\figurewidth}
\setlength{\figurewidth}{\textwidth}
\addtolength{\figurewidth}{-0.4cm}

% font for program texts
\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
\newcommand{\codefont}{\small\tt}

\usepackage{listings}
\lstset{aboveskip=1.5ex,
        belowskip=1.5ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        xleftmargin=2ex,
        basewidth=0.52em,
        basicstyle=\small\ttfamily}

\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
       }

\lstnewenvironment{curry}{}{}
\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
\newcommand{\listline}{\vrule width0pt depth1.75ex}

% program text in normal text
\newcommand{\code}[1]{\mbox{\codefont #1}}

% program text in normal text with apostrophs
\newcommand{\ccode}[1]{``\code{#1}''}

\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index

\newcommand{\CYS}{Curry\xspace} % name of the Curry system described here
\newcommand{\cyshome}{\mbox{\textit{curryhome}}\xspace} % symbolic installation directory

\begin{document}
\sloppy

\input{manual.tex}

% Bibliography
\begin{thebibliography}{10}

\bibitem{AntoyHanus05LOPSTR}
S.~Antoy and M.~Hanus.
\newblock Declarative programming with function patterns.
\newblock In {\em Proceedings of the International Symposium on Logic-based
  Program Synthesis and Transformation (LOPSTR'05)}, pages 6--22. Springer LNCS
  3901, 2005.

\bibitem{AntoyHanus09}
S.~Antoy and M.~Hanus.
\newblock Set functions for functional logic programming.
\newblock In {\em Proceedings of the 11th ACM SIGPLAN International Conference
  on Principles and Practice of Declarative Programming (PPDP'09)}, pages
  73--82. ACM Press, 2009.

\bibitem{AntoyHanus12PADL}
S.~Antoy and M.~Hanus.
\newblock Contracts and specifications for functional logic programming.
\newblock In {\em Proc. of the 14th International Symposium on Practical
  Aspects of Declarative Languages (PADL 2012)}, pages 33--47. Springer LNCS
  7149, 2012.

\bibitem{ChristiansenFischer08FLOPS}
J.~Christiansen and S.~Fischer.
\newblock {EasyCheck} - test data for free.
\newblock In {\em Proc. of the 9th International Symposium on Functional and
  Logic Programming (FLOPS 2008)}, pages 322--336. Springer LNCS 4989, 2008.

\bibitem{ClaessenHughes00}
K.~Claessen and J.~Hughes.
\newblock Quickcheck: A lightweight tool for random testing of haskell
  programs.
\newblock In {\em International Conference on Functional Programming
  (ICFP'00)}, pages 268--279. ACM Press, 2000.


\end{thebibliography}

\end{document}
